---
title: "Presentation"
author: "Francesco Vo & Marco Uderzo"
date: "2023-04-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r, include=FALSE}
library(MASS)
library(pROC)
library(class)
library(ggplot2)
library(ggcorrplot)
library(gridExtra)
library(corrplot)
library(correlation)
library(ggm)
library(igraph)
library(tidymodels)
library(naivebayes)
library(ROCR)
library(glmnet)

data <- read.csv("data/heart_data.csv", stringsAsFactors = T)

attach(data)
```

## Project goal and dataset description
https://www.kaggle.com/datasets/fedesoriano/heart-failure-prediction

The dataset consists of medical data from 5 different locations.
This data informs us of the presence of coronary heart disease.

Our main goal is to predict the presence of this condition given the other parameters in the dataset.

This dataset has a number of 12 parameters and presents us 918 observations.

The parameters are:

1. **Age**: age of the patient [years]

2. **Sex**: sex of the patient [M: male, F: female]

3. **ChestPainType**: chest pain type [TA: typical angina, ATA: atypical angina, NAP: non-anginal pain, ASY: asymptomatic]

+ Angina is a type of chest pain caused by reduced blood flow to the heart. Angina is a symptom of coronary heart disease.

4. **RestingBP**: resting blood pressure [mmHg]

5. **Cholesterol**: serum cholesterol [mm/dl]

6. **FastingBS**: fasting blood sugar [1: if FastingBS > 120 mg/dl, 0: otherwise]

+ This measures the blood sugar level after an overnight fast.

7. **RestingECG**: resting electrocardiogram results [Normal: normal, ST: having ST-T wave abnormality (T wave inversions and/or ST elevation or depression of > 0.05 mV), LVH: showing probable or definite left ventricular hypertrophy by Estes' criteria]

8. **MaxHR**: maximum heart rate achieved [numeric value between 60 and 202]

9. **ExerciseAngina**: exercise-induced angina [Y: yes, N: no]

10. **Oldpeak**: oldpeak = ST [numeric value between -2.6 and 6.2]

+ ST depression refers to a finding on an electrocardiogram, wherein the trace in the ST segment is abnormally low below the baseline.
Oldpeak measures the depression of the ST slope.

11. **ST_Slope**: the slope of the peak exercise ST segment [Up: upsloping, Flat: flat, Down: downsloping]

12. **HeartDisease**: output class [1: heart disease, 0: normal]

<br>

## Data visualization and cleaning
In this section we want to visualize our data and clean it. In order to that we have to:

* Visualize the distribution of the target variable

* Visualize the distribution of the predictors

* Deal with missing values

* Deal with outliers

* Find correlations between the parameters

<br>

### Data balance check
```{r, include=TRUE}
prop.table(table(HeartDisease))
```

The dataset is quite balanced.

<br>

### Continuous variables
Now we plot the continuous variables with respect to the target variable.

```{r, include=FALSE}
# continuous variables
colours <- c("#F8766D", "#00BFC4")

# Age
age.plot.1 <- ggplot(data, aes(x=Age, group=HeartDisease,
                             fill=factor(HeartDisease))) +
  geom_density(alpha=0.4) + 
  ggtitle("Age - Density Plot") + xlab("Age") +
  guides(fill = guide_legend(title="Heart disease"))

age.plot.2 <- ggplot(data, aes(x=Age, group=Sex,
                               fill=factor(Sex))) +
  geom_density(alpha=0.4) + 
  ggtitle("Age - Density Plot") + xlab("Age") +
  guides(fill = guide_legend(title="Gender"))

# RestingBP
restingBP.plot <- ggplot(data, aes(x=RestingBP, group=HeartDisease,
                               fill=factor(HeartDisease))) +
  geom_density(alpha=0.4) + 
  ggtitle("RestingBP - Density Plot") + xlab("RestingBP") +
  guides(fill = guide_legend(title="Heart disease"))

# Cholesterol
chol.plot <- ggplot(data, aes(x=Cholesterol, group=HeartDisease,
                                            fill=factor(HeartDisease))) +
  geom_density(alpha=0.4) + 
  ggtitle("Cholesterol - Density Plot") + xlab("Cholesterol") +
  guides(fill = guide_legend(title="Heart disease"))

# MaxHR
maxHR.plot <- ggplot(data, aes(x=MaxHR, group=HeartDisease,
                               fill=factor(HeartDisease))) +
  geom_density(alpha=0.4) + 
  ggtitle("MaxHR - Density Plot") + xlab("MaxHR") +
  guides(fill = guide_legend(title="Heart disease"))

# Oldpeak
oldpeak.plot <- ggplot(data, aes(x=Oldpeak, group=HeartDisease,
                               fill=factor(HeartDisease))) +
  geom_density(alpha=0.4) + 
  ggtitle("Oldpeak - Density Plot") + xlab("Oldpeak") +
  guides(fill = guide_legend(title="Heart disease"))

# FastingBS
fastingBS.plot <- ggplot(data, aes(x=FastingBS, group=HeartDisease,
                                   fill=factor(HeartDisease))) +
  geom_bar(alpha=0.5, position="dodge") +
  guides(fill = guide_legend(title="Heart disease"))

```

```{r, include=TRUE}
chol.plot
oldpeak.plot
fastingBS.plot
```

```{r, include=TRUE}
age.box.1 <- boxplot(Age ~ HeartDisease, col=colours)
restingBP.box <- boxplot(RestingBP ~ HeartDisease, col=colours)
maxHR.box <- boxplot(MaxHR ~ HeartDisease, col=colours)
```

<br>

### Categorical variables
```{r, include=FALSE}
# Sex
sex.plot <- ggplot(data, aes(x=Sex, group=HeartDisease,
                                   fill=factor(HeartDisease))) +
  geom_bar(alpha=0.5, position="dodge") +
  guides(fill = guide_legend(title="Heart disease"))

# ChestPainType
cpt.plot <- ggplot(data, aes(x=ChestPainType, group=HeartDisease,
                                   fill=factor(HeartDisease))) +
  geom_bar(alpha=0.5, position="dodge") +
  guides(fill = guide_legend(title="Heart disease"))

# RestingECG
restingECG.plot <- ggplot(data, aes(x=RestingECG, group=HeartDisease,
                                   fill=factor(HeartDisease))) +
  geom_bar(alpha=0.5, position="dodge") +
  guides(fill = guide_legend(title="Heart disease"))

# ExerciseAngina
exAn.plot <- ggplot(data, aes(x=ExerciseAngina, group=HeartDisease,
                                   fill=factor(HeartDisease))) +
  geom_bar(alpha=0.5, position="dodge") +
  guides(fill = guide_legend(title="Heart disease"))

# ST_Slope
st.plot <- ggplot(data, aes(x=ST_Slope, group=HeartDisease,
                                   fill=factor(HeartDisease))) +
  geom_bar(alpha=0.5, position="dodge") +
  guides(fill = guide_legend(title="Heart disease"))

```

```{r, include=TRUE}
sex.plot
cpt.plot
restingECG.plot
exAn.plot
st.plot
```

### Missing values
We have seen from the plots that there many variables in Cholesterol that have value equal to 0. We assume that those variables are not plausible and we change them.

Since we don't want to delete data because we don't have many, in this case we replace the missing values with random values that follow a normal distribution with mean and standard deviation equal to that of the data (we don't want to influence too much).
``` {r, include=TRUE}
data$Cholesterol[data$Cholesterol == 0] <- round(rnorm(length(data$Cholesterol[data$Cholesterol == 0]),
                                                 mean(Cholesterol), sd(Cholesterol)),0)
```

<br>

### Outliers
We decided not to deal with outliers since we don't have much data and we don't want to influence too much the results of our classifiers.

<br>

### Correlations
```{r, include=TRUE}
model.matrix(~0+., data=data) %>%
  cor(use="pairwise.complete.obs") %>%
  ggcorrplot(show.diag=FALSE, lab=T, lab_size=1.5, tl.cex=5)
```
